---
title: "Figure 5"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r message = FALSE}
library(readr)
library(dplyr)
library(purrr)
```

## Capacity formula

To make things simpler, let's consider only the ingredients of capacity that we
have studied here as a function of age $a$:

$$
C(a) \propto  \frac{p(a) \times i(a)}{g(a)} e^{-\frac{g(a)}{m(a)}}
$$
Everything else will be a multiplicative constant. We don't care too much about
the exact values of such a constant because we will express everything relative
to the 4-day-old group. Futhermore, in the results, we will learn from
experiment 3 that $m(a)$ does not change with age. So, for figure 5 we will be
able to simply consider the following formula:

$$
C(a) \propto  \frac{p(a) \times i(a)}{g(a)} e^{-\frac{g(a)}{m}}
$$
Sporozoite prevalences $p(a)$ and sporozoite intensities (given infection)
$i(a)$ are informed by experiment 1 and mosquitoes survival rates $g(a)$ are
informed by experiment 2. We will assume that the vector competence and survival
are independent from each other. We'll assume that $p(a)$ follows a beta
distribution, the parameters of which will be estimated from the binomial
estimate and its confidence interval from the presence/absence raw data. Values
for $p(a)$ will then be randomly drawed from this beta distribution. Given that
results from experiment 1 show that the distributions of sporozoites intensities
do not look like any clear distribution, it will probably be better to perform
non-parametric bootstrapping of the data in order to generate values of
sporozoites intensities. As for the mosquito survival rates, we will take the
estimates from the mixed-effect model of experiment 2.

## Data from experiment 1

Loading and preparing the data:

```{r message = FALSE}
fig2 <- "data/fig2_sporozoite_data.txt" |>
  read_delim() |> 
  filter(age_class != "8-day-old") |> 
  mutate(intensity = 2^{-Ct})
```

The names of the age classes:

```{r}
age_names <- sort(unique(fig2$age_class))
```

### Sporozoite prevalences

A function that converts confidence intervals boundary values of a proportion
into shape parameters of a beta distribution:

```{r}
ci2pars <- function(ci, f, pars, p = .95) {
  p <- (1 - p) / 2
  
  f_par <- function(par1, par2) {
    c(f(p, par1, par2), f(1 - p, par1, par2))
  }

  f_fit <- function(pars) {
    sum((f_par(pars[1], pars[2]) - ci)^2)
  }
  
  optim(pars, f_fit)$par
}
```

The shape parameters of the beta distribution of the mean sporozoites
prevalences ($p(a)$):

```{r warning = FALSE}
p_par <- fig2 |> 
  group_by(age_class) |> 
  summarize(x   = sum(positive), n = length(positive)) |> 
  mutate(prop   = map2(x, n, prop.test),
         ci     = map(prop, magrittr::extract2, "conf.int"),
         shapes = map(ci, ci2pars, qbeta, 1:2)) |> 
  pull(shapes) |> 
  setNames(age_names)
```

### Sporozoite intensities

The data on sporozoites intensities, given infection:

```{r}
i <- fig2 |> 
  filter(positive > 0) |> 
  group_by(age_class) |> 
  group_split() |> 
  map(pull, intensity) |> 
  setNames(age_names)
```

```{r}
hist(log(i$`4-day-old`), n = 50)
hist(log(i$`12-day-old`), n = 50)
```

## Experiment 2

```{r message = FALSE}
fig3 <- "data/fig3_survival_data.txt" |>
  read_delim() |> 
  filter(age_class != "8-day-old")
```

```{r}
fig3 |> 
  filter(age_class == "4-day-old") |> 
  group_by(daysPI) |> 
  tally() |> 
  mutate(d = cumsum(n)) |> 
  with(plot(daysPI, d))
```

```{r}
age4 <- fig3 |> 
  filter(age_class == "4-day-old") |> 
  pull(daysPI)
```

```{r}
inf_status <- sort(unique(fig3$infection_status))

age12 <- fig3 |> 
  filter(age_class == "12-day-old") |> 
  group_by(infection_status) |> 
  group_split() |> 
  map(pull, daysPI) |> 
  setNames(inf_status)
```

## Experiment 3

```{r message = FALSE}
fig4 <- "data/fig4_EIP_data.txt" |>
  read_delim() |> 
  select(dpi, spz) |> 
  na.exclude() |> 
  mutate_at("spz", as.logical)
```


```{r warning = FALSE}
mod <- glm(spz ~ dpi, binomial, fig4)
xs <- seq(min(fig4$dpi), max(fig4$dpi), le = 1000)
ys <- predict(mod, data.frame(dpi = xs), se.fit = TRUE)
fit <- ys$fit
invtransf <- family(mod)$linkinv
low <- invtransf(ys$fit - 1.96 * ys$se.fit)
upp <- invtransf(ys$fit + 1.96 * ys$se.fit)
fit <- invtransf(fit)

plot(xs, fit, type = "l")
lines(xs, low, lty = 2)
lines(xs, upp, lty = 2)
abline(h = .5)

ci <- tibble(xs, low, upp)

m <- ci2pars(c(filter(ci, .5 < upp)$xs[1], filter(ci, .5 < low)$xs[1]), qgamma, 1:2)
```

## Simulating data

```{r}
simulate <- function(p, i, g, N = 10000) {
  p_val <- do.call(rbeta, c(list(N), unname(p)))
  i_val <- sample(i, N, replace = TRUE)
  g_val <- 1 / sample(g, N, replace = TRUE)
  m_val <- do.call(rgamma, c(list(N), m))
  p_val * i_val * exp(-g_val) / g_val
}
```

```{r}
sim_age4           <- simulate(p_par$`4-day-old`,   i$`4-day-old`, age4)
sim_age12_ExpInf   <- simulate(p_par$`12-day-old`, i$`12-day-old`, age12$`Exposed-infected`)
sim_age12_ExpNoinf <- simulate(p_par$`12-day-old`, i$`12-day-old`, age12$`Exposed-uninfected`)
sim_age12_Uninf    <- simulate(p_par$`12-day-old`, i$`12-day-old`, age12$`Uninfected control`)
```

```{r}
hist(log(sim_age4))
hist(log(sim_age12_ExpInf))
hist(log(sim_age12_ExpNoinf))
hist(log(sim_age12_Uninf))
```

Checking that stationarity has been reached:

```{r}
walk(list(sim_age4, sim_age12_ExpInf, sim_age12_ExpNoinf, sim_age12_Uninf),
     function(x, ...) plot(cummean(x), type = "l", ...))
```

